Synthesising Recursive Functions for First-Order Model Counting

### 2023-02-14 doc/paper/supplementary_material.tex
Move things around

% TODO (at the very end)
% grammarly the whole thing
% read the formatting instructions and their LaTeX source (email font?)
% make sure there are no overflow errors
%  make sure the result table (as well as other floating objects) are on the right pages

@@

\maketitle

\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in finite-domain first-order logic. Despite
  being around for more than a decade, practical FOMC algorithms are still
  unable to compute functions as simple as a factorial. We argue that the
  capabilities of FOMC algorithms to date are limited by their inability to
  express arbitrary recursive computations. To enable arbitrary recursion, we
  relax the restrictions that typically accompany domain recursion and
  generalise circuits used to express a solution to an FOMC problem to graphs
  that may contain cycles. To this end, we enhance the most well-established
  (weighted) FOMC algorithm \textsc{ForcLift} with new compilation rules and an
  algorithm to check whether a recursive call is feasible. These improvements
  allow the algorithm to construct efficient solutions to counting fundamental
  structures such as injections and bijections. We end with a few conjectures on
  what classes of instances could be liftable as a result.
\end{abstract}

\section{Introduction}\label{sec:introduction}

% 1. What is the problem?
% 2. Why is it interesting and important?

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. Lifted inference
algorithms are typically used on probabilistic models such as probabilistic
programming
languages~\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17},
Markov logic
networks~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical~\cite{DBLP:journals/ml/KimmigMG15} and statistical
relational~\cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC has received more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}. There is also a recent line
of work that uses FOMC-based methods to count solutions to constraint
satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23}.

% 3. Why is it hard? (E.g., why do naive approaches fail?)
% complexity/liftability -> murky boundary

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two seminal results, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}, and
\Ctwo{} extended with tree axioms~\cite{DBLP:conf/kr/BremenK21} and linear order
axiom~\cite{DBLP:journals/corr/abs-2211-01164}. On the empirical front, there
are several implementations of exact WFOMC algorithms:
\textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16},
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21},
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{Alchemy}, \textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20},
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12},
\textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}, and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}.

% Magician: an approach to improve the efficiency of sampling and local search
% based MLN inference algorithms by utilising \#CSP algorithms. Tuffy: grounding
% and stochastic local search.

% 4. Why hasn't it been solved before? (Or, what's wrong with previous proposed
% solutions? How does mine differ?)
% simple unliftable instances -> recursion

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is some previous work
  showing how FOMC can compute
  factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
  One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
  not directly correspond to any algorithm.
  Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
  liftable but contains no experimental results. The
  third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
  with an algorithm, which is not publicly available.} We claim that the
capabilities of (W)FOMC algorithms can be significantly extended by empowering
them with the ability to construct recursive solutions. The topic of recursion
in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

% 5. What are the key components of my approach and results? Also include any
% specific limitations.

% 5.1) An overview of how ForcLift/Crane work.

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy search, \textsc{Crane} also supports a hybrid search algorithm
that applies some rules greedily and some using breadth-first
search.\footnote{In the current implementation, the rules that are applied in a
  non-greedy fashion are: atom counting, inclusion-exclusion, independent
  partial groundings, Shannon decomposition, shattering, and two of our new
  rules introduced in \cref{sec:dr,sec:ref}---see previous
  work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for more information about the
  rules.} This alternative was introduced since there is no reason to expect
greedy search to be optimal. Second, the product of compilation is not directly
evaluated but rather interpreted as a collection of functions on domain sizes.
Hence, our approach is reminiscent of previous work on lifted inference via
compilation to C++ programs~\cite{DBLP:conf/kr/KazemiP16} and the broader area
of functional
synthesis~\cite{DBLP:conf/cav/GoliaRM20,DBLP:conf/pldi/KuncakMPS10,sanathanan1963transfer}.

% similarly to the other domains, there are many functions that fit the
% specification, i.e., can correctly count the number of models as a function of
% domain size(s). Whereas in [...] one typically wants to select the smallest
% definition of a function, we are primarily concerned with its worst-case
% complexity as measured by the number of arithmetic operations needed for an
% arbitrary function call
% \begin{itemize}
%   \item Boolean functions~\cite{DBLP:conf/cav/GoliaRM20}
%   \item dynamical systems~\cite{sanathanan1963transfer}
%   \item program fragments~\cite{DBLP:conf/pldi/KuncakMPS10}
% \end{itemize}

% \footnote{There is also an intermediate stage called \emph{smoothing} that takes
%   place between compilation and evaluation. As our changes to smoothing are
%   rather elementary, we do not discuss them to not distract the reader from the
%   main contributions of this paper.}

% 5.2) The main idea: cycles that represent recursive calls.

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits, where cycles
represent recursive calls. Suppose the input formula $\phi$ depends on a domain
of size $n \in \mathbb{N}_{0}$. \emph{Generalised domain recursion} (GDR)---one
of the new compilation rules---transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another compilation rule recognizes that the
resulting formula is just like the input formula $\phi$ but with a different
domain. This observation allows us to add a cycle-forming edge to the graph,
which can be interpreted as function $f$ relying on $f(n-1)$ to compute $f(n)$.

% 5.3) contributions, references to sections, shortcomings

We begin by defining the representation used for sentences in FOL in
\cref{sec:recprelims}. Then, in \cref{sec:methods}, we define the graphs that
replace circuits in representing a solution to a (W)FOMC problem.
\Cref{sec:rules} introduces the new compilation rules. In \cref{sec:interpret},
we discuss how a graph can be interpreted as a collection of (potentially
recursive) functions. In \cref{sec:results}, we compare \textsc{ForcLift} and
\textsc{Crane} on a range of function-counting problems. We show that
\textsc{Crane} performs as well as \textsc{ForcLift} on the instances that were
already solvable by \textsc{ForcLift} but is also able to handle most of the
instances that \textsc{ForcLift} fails on. Finally, in \cref{sec:conclusion}, we
conclude by outlining some conjectures and directions for future work.

\section{Preliminaries}\label{sec:recprelims}

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can
translate sentences in a variant of function-free many-sorted FOL with equality
to this internal format. We use lowercase Latin letters for predicates (e.g.,
$\texttt{p}$) and constants (e.g., $x$), uppercase Latin letters for variables
(e.g., $X$), and uppercase Greek letters for domains (e.g., $\Delta$). Sometimes
we write predicate $\texttt{p}$ as $\texttt{p}/n$, where $n \in \mathbb{N}^{+}$
is the \emph{arity} of $\texttt{p}$. An \emph{atom} is
$\texttt{p}(t_1, \dots, t_n)$ for some predicate $\texttt{p}/n$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg \texttt{p}(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all
relevant (finite) domains. Initially, $\mathcal{D}$ contains all domains
mentioned by the input formula. During compilation, new domains are added to
$\mathcal{D}$ by some of the compilation rules. Each such new domain is
interpreted as a subset of some other domain in $\mathcal{D}$.

\begin{definition}[Constraint]\label{def:constraint}
  An \emph{(inequality) constraint} is a pair $(a, b)$, where $a$ is a variable,
  and $b$ is either a variable or a constant.
\end{definition}

\begin{definition}[Clause]\label{def:clause}
  A \emph{clause}\footnote{Van den Broeck et
    al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} refer to clauses as c-clauses.}
  is a triple $c = (L, C, \delta_c)$, where $L$ is a set of literals, $C$ is a
  set of constraints, and $\delta_c$ is the domain map of $c$. Let $\Vars$ be
  the function that maps clauses and sets of either literals or constraints to
  the set of variables contained within. In particular,
  $\Vars(c) \coloneqq \Vars(L) \cup \Vars(C)$. \emph{Domain map}
  $\delta_{c}\colon \Vars(c) \to \mathcal{D}$ is a function that maps all
  variables in $c$ to their domains such that (s.t.) if $(X, Y) \in C$ for some
  variables $X$ and $Y$, then $\delta_c(X) = \delta_c(Y)$. For convenience, we
  sometimes write $\delta_c$ for the domain map of $c$ without unpacking $c$
  into its three constituents.
\end{definition}

\paragraph{Notation.}
For any set $S$, let $2^{S}$ denote its power set and
$S^{\ast} \coloneqq \bigcup_{i=0}^{\infty} S^{i}$ the set of tuples of any
finite length of elements of $S$. For any $n \in \mathbb{N}_{0}$, let
$[n] \coloneqq \{\, 1, \dots, n \,\}$. For example, $[0] = \emptyset$, and
$[2] = \{\, 1, 2 \,\}$. We write $\langle\rangle$ for an empty list,
$\langle x \rangle$ for a list with one element $x$, $|l|$ for the length of
list $l$, and $\pfun$ to denote partial functions. Let $S$ be a set of
constraints or literals, $V$ a set of variables, and $x$ a variable or a
constant. We write $S[x/V]$ to denote $S$ with all occurrences of all variables
in $V$ replaced with $x$. Let $\Doms$ (respectively, $\Preds$) be the function
that maps any clause or formula to the set of domains (respectively, predicates)
used within, and let $\dom(f)$ denote the domain of function $f$.

\begin{definition}[Formula]\label{def:formula}
  A \emph{formula} (called a c-theory by Van den Broeck et
  al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}) is a set of clauses s.t.\ all
  constraints and atoms `type check' with respect to domains. Each formula
  $\phi$ has a map $\pi_{\phi}\colon \Preds(\phi) \to {\Doms(\phi)}^{\ast}$ such
  that, for each predicate $\texttt{p}/n \in \Preds(\phi)$, we have that
  $\pi_{\phi}(\texttt{p}) \in {\Doms(\phi)}^{n}$.\footnote{Similarly to
    variables and predicates, constants are also mapped to domains, although we
    keep this mapping implicit.}
\end{definition}

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $\texttt{p}/2$, variables $X$, $Y$, $Z$, and domains
  $\Gamma$ and $\Delta$. Based on $\delta_{c_{1}}$ and $\delta_{c_{2}}$, we can
  infer that $\pi_{\phi} = \{\, \texttt{p} \mapsto (\Gamma, \Delta) \,\}$. All
  variables that occur as the first argument to $\texttt{p}$ are in $\Gamma$,
  and, likewise, all variables that occur as the second argument to $\texttt{p}$
  are in $\Delta$. Therefore, $\phi$ `type checks' as a valid formula.
\end{example}

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
First, we decouple variable-to-domain assignments from constraints and move them
to a separate function $\delta_{c}$ in \cref{def:clause}. Second, while Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} allow for equality
constraints and constraints of the form $X \not\in \Delta$ for some variable $X$
and domain $\Delta$, we exclude such constraints simply because they are not
needed.

One can read a formula in our format as a sentence in FOL\@. All variables in a
clause are implicitly universally quantified (but note that variables are never
shared among clauses), and all clauses in a formula are implicitly linked by a
conjunction. Thus, formula $\phi$ from \cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(Z, Y)).
    \end{multlined}
\end{align*}

Once domains are mapped to finite sets and constants to elements in those sets,
a formula can be viewed as a set of conditions that the predicates (interpreted
as relations) have to satisfy. Hence, FOMC is the problem of counting the number
of combinations of relations that satisfy these conditions.

\begin{example}\label{example:simple}
  Let $\Delta$ be a domain of size $n \in \mathbb{N}_{0}$. The model count of
  $\forall X \in \Delta\text{. } \texttt{p}(X) \lor \texttt{q}(X)$ is then
  $3^{n}$. Intuitively, since both predicates are of arity one, they can be
  interpreted as subsets of $\Delta$. Thus, the formula says that each element
  of $\Delta$ has to be included in $\texttt{p}$ or in $\texttt{q}$ or in both,
  i.e., there are three possibilities.
\end{example}

\begin{example}\label{example:smokers}
  Consider a variant of the well-known `friends and smokers' example
  $\forall X, Y \in \Delta\text{.
  } \smokes(X) \land \friends(X, Y) \Rightarrow \smokes(Y)$. Letting
  $n \coloneqq |\Delta|$ as in \cref{example:simple}, the model count can be
  expressed as
  $\sum_{k=0}^{n} \binom{n}{k}2^{n^{2} - k(n-k)}$~\cite{DBLP:conf/kr/BroeckMD14}.
\end{example}

To formalise the idea of a model, let
$\sigma\colon \mathcal{D} \to \mathbb{N}_{0}$ be the \emph{domain size function}
that maps each domain to a non-negative integer.

\begin{definition}\label{def:model}
  Let $\phi$ be a formula and $\sigma$ a domain size function. A \emph{model} of
  $(\phi, \sigma)$ is a map
  $\mathfrak{M}\colon \Preds(\phi) \to 2^{\mathbb{N}_{0}^{\ast}}$ such that the
  following two conditions are satisfied.
  \begin{enumerate}
    \item Let $\texttt{p}/n \in \Preds(\phi)$ be a predicate and let
          $\pi_{\phi}(\texttt{p}) = {(d_{i})}_{i=1}^{n}$ for some domains
          $d_{i} \in \Doms(\phi)$. Then
          \begin{equation}\label{eq:subset}
            \mathfrak{M}(\texttt{p}) \subseteq \prod_{i=1}^{n}[\sigma(d_{i})].\footnote{For simplicity, \cref{def:model} ignores constants. To include constants, one would replace $[\sigma(d_{i})]$ with a set that contains all constants associated with domain $d_{i}$, extended with enough new elements to make its cardinality $\sigma(d_{i})$.}
          \end{equation}
    \item As a collection of relations, $\mathfrak{M}$ satisfies $\phi$.
  \end{enumerate}
\end{definition}

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let
  $\sigma(\Gamma) = \sigma(\Delta) = 2$. Then the right-hand side of
  \cref{eq:subset} becomes
  \[
    [\sigma(\Gamma)] \times [\sigma(\Delta)] = [2] \times [2] = \{\,(1, 1), (1, 2), (2, 1), (2, 2)\,\}
  \]
  There are $2^{4} = 16$ possible relations between $\Gamma$ and $\Delta$. Let
  us count how many of them satisfy the conditions imposed on predicate
  $\texttt{p}$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $(\phi, \sigma)$ is 7. Incidentally, it
  counts partial injections. We will continue to use the problem of counting
  partial injections (and the formula from \cref{example:first} specifically) as
  the main running example throughout the paper.
\end{example}

\section{From Circuits to Graphs}\label{sec:methods}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree={s sep=10mm, sn edges}
    [$\bigwedge_{x \in \Delta}$,ellipse,draw,label=right:{\color{blue}{$3^{n}$}}
    [$\lor$,ellipse,draw,label=right:{\color{blue}{$2 + 1 = 3$}}
    [$\land$,ellipse,draw,label=left:{\color{blue}{$2 \times 1 = 2$}}
    [$p(x) \lor \neg p(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{2}}]
    [$\land$,ellipse,draw,label={[label distance=0cm,text=blue]87:$1 \times 1 = 1$}
    [$q(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    [$\top$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    ]
    ]
    [$\land$,ellipse,draw,label=right:{\color{blue}{$1 \times 1 = 1$}}
    [$\neg q(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    [$p(x)$,rectangle,draw,fill=gray!25,label=below:{\color{blue}{1}}]
    ]
    ]
    ]
  \end{forest}
  \caption{A circuit produced by \textsc{ForcLift} for \cref{example:simple}.
    Values and computations in blue (on the outside of each node) show how a
    bottom-up evaluation of the circuit computes the model
    count.}\label{fig:simplecircuit}
\end{figure}

Darwiche~\shortcite{DBLP:journals/jancl/Darwiche01} introduced
\emph{deterministic decomposable negation normal form} (d-DNNF) circuits for
propositional knowledge compilation and showed that the model count of a
propositional formula can be computed in time linear in the size of the circuit.
Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} generalised them
to FOL via \emph{first-order d-DNNF (FO d-DNNF) circuits}. FO d-DNNF circuits
(hereafter simply called \emph{circuits}) are directed acyclic graphs with nodes
corresponding to formulas in FOL---see \cref{fig:simplecircuit} for an example.
The following types of nodes are supported by \textsc{ForcLift}: caching
($\Reff$), contradiction ($\bot$), tautology ($\top$), decomposable conjunction
($\land$), decomposable set-conjunction ($\bigwedge$), deterministic disjunction
($\lor$), deterministic set-disjunction ($\bigvee$), domain recursion,
grounding, inclusion-exclusion, smoothing, and unit clause. We refer the reader
to previous work~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11} for
more information about node types and their interpretations for computing the
(W)FOMC\@.

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$,ellipse,draw,name=gdr
    [$\bigvee_{\Delta^\top \subseteq \Delta}$,ellipse,draw
    [$\CR_{\Gamma \mapsto \Gamma'}$,ellipse,draw
    [$\land$,ellipse,draw
    [$\bot$,rectangle,draw,fill=gray!25]
    [$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$,ellipse,draw,name=ref]
    ]
    ]
    ]
    ]
    \draw[-Latex,bend right=45] (ref) to (gdr);
  \end{forest}
  \caption{A simplified version of the FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Here we
    omit some parameters as well as nodes whose only arithmetic effect is
    multiplication by one.}\label{fig:examplefcg}
\end{figure}

In this paper, we introduce \emph{FO d-DNNF computational graphs} (FCGs) that
generalise circuits by dispensing with acyclicity. Formally, an FCG is a (weakly
connected) directed graph with a single source, node labels, and ordered
outgoing edges. Node labels consist of two parts: the \emph{type} and the
\emph{parameters}. The type of a node determines its out-degree. We make the
following changes and additions to the node types already supported by
\textsc{ForcLift}. First, we introduce a new node type for constraint removal
($\CR$). Second, we replace domain recursion with generalised domain recursion
($\GDR$). And third, for reasoning about partially-constructed FCGs, we write
$\star$ for a placeholder type that is yet to be replaced. Similarly to Van den
Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an
FCG that has a node with the label $T_p$ (i.e., type $T$ and parameter(s) $p$)
and $\star$'s as all of its direct successors. See \cref{fig:examplefcg} for an
example FCG\@. Its source node has out-degree 1, label
$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, s.t. $|L|$ is equal to the number
of $\star$'s in $G$. $L$ contains formulas that still need to be compiled. Once
a formula is compiled, it replaces one of the $\star$'s in $G$ according to a
set order. We say that an FCG is \emph{complete} (i.e., it represents a
\emph{complete solution}) if it has no $\star$'s. Similarly, a chip is complete
if its FCG is complete (or, equivalently, the list of formulas is empty).

\section{New Compilation Rules}\label{sec:rules}

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one
of them, in an attempt to not miss a solution, \textsc{Crane} returns them all.
In particular, if a rule returns an empty set, then that rule does not apply to
the formula.

\subsection{Generalised Domain Recursion}\label{sec:dr}

% Main idea
The main idea behind domain recursion (both the original version by Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11} and the one presented here) is as
follows. Let $\Omega \in \mathcal{D}$ be a domain. Assuming that
$\Omega \ne \emptyset$, pick some $x \in \Omega$. Then, for every variable
$X \in \Omega$ that occurs in a literal, consider two possibilities: $X = x$ and
$X \ne x$.

\begin{example}\label{example:dr}
  Let $\phi$ be a formula with a single clause
  \begin{multline*}
    (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}).
  \end{multline*}
  Then we can introduce constant $x \in \Gamma$ and rewrite $\phi$ as
  $\phi' = \{\, c_{1}, c_{2} \,\}$, where
  \begin{align*}
    c_{1} &= \begin{multlined}[t]
      (\{\, \neg \texttt{p}(x, Y), \neg \texttt{p}(x, Z) \,\}, \{\, (Y, Z) \,\}, \\
      \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
      \end{multlined}\\
    c_{2} &= \begin{multlined}[t]
      (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
      \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Delta \,\}),
      \end{multlined}
  \end{align*}
  and $\Gamma' = \Gamma \setminus \{\, x \,\}$.
\end{example}

% Original domain recursion
Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} imposes stringent conditions
on the input formula to ensure that the expanded version of the formula (as in
\cref{example:dr}) can be handled efficiently. For instance,
\cref{example:first} cannot be handled by \textsc{ForcLift} because there is no
root binding class, i.e., the two root variables belong to different equivalence
classes with respect to the binding relationship. The clauses in this expanded
formula are then partitioned into three parts based on whether the
transformation introduced constants or constraints or both. The aforementioned
conditions ensure that these parts can be treated independently.

>>>>

% TODO (at the very end)
% put the appendix in a separate file
% grammarly the whole thing
% read the formatting instructions and their LaTeX source (email font?)
% make sure there are no overflow errors
% clean up unnecessary things from the header (e.g., packages and definitions)
%  make sure the result table (as well as other floating objects) are on the right pages

@@

\appendix
\section{Solutions Found by \textsc{Crane}}\label{app:solutions}

In this appendix, we list the exact function definitions produced by
\textsc{Crane} for all of the problem instances in Section~6 both before and
after algebraic simplification (excluding multiplications by one). The
correctness of all of them has been checked by identifying suitable base cases
and verifying the numerical answers across a range of domain sizes.

\begin{enumerate}
  \item $\Theta(m)$ solution for counting $\Gamma \to \Gamma$ functions:
  \[
    f(m) = {\left(-1 + \sum_{l=0}^{m} \binom{m}{l} [l < 2]\right)}^{m} = m^{m}.
  \]
  \item $\Theta(m^3 + n^3)$ solution for counting $\Gamma \to \Delta$
  surjections:
  \begin{align*}
    f(m, n) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k}\\
               &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
    ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k} {(k+1)}^{l}.
  \end{align*}
  \item $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ surjections:
  \begin{align*}
    f(m) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k}\\
            &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
    ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k} {(k+1)}^{l}.
  \end{align*}
  \item $\Theta(mn)$ solution for counting $\Gamma \to \Delta$ injections and
  partial injections (with different base cases):
  \begin{align*}
    f(m, n) ={}& \sum_{l=0}^m \binom{m}{l} [l<2] f(m-l, n-1)\\
    ={}& f(m, n-1) + mf(m-1, n-1).
  \end{align*}
  \item $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ injections:
  \begin{align*}
    f(m) ={}& \sum_{l=0}^{m} \binom{m}{l} {(-1)}^{m-l} g(m, l); \\
    g(m, l) ={}& \sum_{k=0}^{l} \binom{l}{k} [k < 2] g(m - 1, l - k)\\
    ={}& g(m - 1, l) + lg(m - 1, l - 1).
  \end{align*}
  % \item $\Theta(m^{2})$ solution for counting $\Gamma \to \Gamma$ partial
  %       injections:
  % \begin{align*}
  %   f(m) ={}& f(m - 1) + \sum_{l = 0}^{m} \binom{m}{l} \sum_{k = 0}^{m} \binom{m}{k}\\
  %           &[1 \le l < 3] [k < 2] 2^{m - l} g(m - l, m - k - 1)\\
  %   ={}& f(m - 1) + m2^{m-1}g(m-1, m-1)\\
  %   +{}& m^{2}2^{m-1}g(m-1, m-2)\\
  %   +{}& \binom{m}{2}2^{m-2}g(m-2, m-1) \\
  %   +{}& \binom{m}{2}m2^{m-2}g(m-2, m-2)\\
  %   g(i, j) ={}& \sum_{h = 0}^{j} \binom{j}{h} [h < 2] g(i - 1, j - h)\\
  %   ={}& g(i-1, j) + jg(i-1, j-1).
  %   \end{align*}
  \item $\Theta(m)$ solution for counting $\Gamma \to \Delta$ bijections:
  \[
    f(m, n) = mf(m-1, n-1).
  \]
\end{enumerate}
